---
title: "1_compute_flow_signatures"
author: "Anya Leenman"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    self_contained: true
---

Code for the paper "Accelerated river mobility linked to water discharge variability" by Leenman et al., submitted June 2024.

This code is to quality-check flow gauge records and compute flow signatures describing discharge variability as well as  bulk flow metrics e.g. mean/median discharge.

We do not have permission to publish the raw time-series data we assembled for this paper from various government agencies, although links to all these agencies are given in the manuscript and supporting material of our paper.

To make a reproducible example of how we computed flow signatures, here we use time-series data accessible in R with the "tidyhydat" package.

## Housekeeping:

```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
library(dplyr)
library(lubridate)
library(tidyhydat)
# download_hydat() # only need to run once

# parameters to set:
#--------
start_year <- 1900 # what year should time series begin (if we need to truncate)
completeness_thresh <- 0.7 # how complete should data be to be processed?
min_len <- 5 # what is minimum record length, in years?
cutoff_month <- 7 #cutoff for "middle" of year - each landsat mosaic from Greenberg et al. 2024 is Jan-Dec, so assign them "1 July" as their date.

outdir <- paste0('./data/SM_data/') # where should outputs be saved?


```

## Data imports

Import

- flow gauges to be used
- time series for each gauge

```{r imports}

# gauges to be used
Q_locations <- c('08HA010', '08LG010', '08LG065',  '05BH004') # a few gauges w/ diff. Q variability.

# time series import
Q_series <- lapply(1:length(Q_locations), FUN = function(i){
  
  f <- hy_daily_flows(station_number = Q_locations[i]) %>%   # read in data from tidyhydat
    select(Date, Value) %>% # select columns of interest
    rename_with(~ Q_locations[i], starts_with('Val')) %>% # rename so discharge column named after site 
    rename(date = Date) %>% 
      mutate(date = ymd(date)) # convert to date format using lubridate (if needed - tidyhydat dates already imported as 'Date' format)

  
  # check record length and completeness (quality)
  #--------------
  # first, check days with no data HAVE been assigned a row:
  f <- f %>%
    mutate(timegaps = date - lag(date) == 1) # extra column to identify gaps; will give 'FALSE' if difference between two dates != 1, i.e. there is a gap.
  
  #---- check for missing data -----
  if(FALSE %in% f$timegaps){
    print(paste0('Timegap in station ', Q_locations[i]))
    new_times <- data.frame(date = seq.Date(min(f$date, na.rm = T), max(f$date, na.rm = T), by="day"))
    f <- full_join(new_times, f, by = 'date')
  }
  
  # remove timegaps column now
  f <- f %>% 
    select(-timegaps)
  
  f_len <- max(year(f$date), na.rm = T) - min(year(f$date), na.rm = T)
  f_qual <- 1 - sum(is.na(f[2]))/nrow(f)
  
  # set data to NA if needed
  if(f_len < min_len | f_qual < completeness_thresh){
    print(paste0('Qual check failed for station ', Q_locations[i]))
    f[2] <- as.numeric(NA)
  } 
  
  return(f)
  
}) %>% 
  purrr::reduce(full_join, by = "date") # join all imported files by 'date', so we end with a dataframe w' a column for date and a discharge column for each site.

Q_series <- Q_series %>% 
  select(!where(~all(is.na(.)))) %>% # remove columns w/ only NA (i.e. didn't pass quality check above)
  arrange(date) %>% 
  filter(date >= ymd(paste0(start_year, '-01-01'))) %>%  # date cutoff, if needed.
  mutate(img_yr = ifelse(month(date) < cutoff_month, # assign "2015-2016" to all dates from 2015-07-01 to 2016-06-30, inclusive.
                         paste0(year(date) - 1, '-', year(date)), 
                         paste0(year(date), '-', year(date) + 1))) %>% 
  relocate(img_yr, .after = date) 


```

## Qual-check

Plot each time series to manually inspect for evidence of major non-stationarity e.g. dam closure.

```{r manual_qualcheck}

for(i in 3:length(Q_series)){
  plot(Q_series$date, Q_series[i] %>% pull(), type = 'l',
       xlim = )
}
```


## Compute flow signatures

CV = Coefficient of variation.

RBI = Richards-Baker Index: https://doi.org/10.1111/j.1752-1688.2004.tb01046.x 

DVIa = annual discharge variability index: https://doi.org/10.1016/j.earscirev.2019.102977

DVIc = cumulative discharge variability index: modified from https://doi.org/10.1016/j.earscirev.2019.102977

Note modified DVIc: instead of using wettest/driest *month* in whole record, we used the 95th and 5th percentile of the *daily* discharge record to compute this metric.


```{r flowsigs}

# --- Function to compute coefficient of variation (CV) of discharge -----
compute_CV <- function(fl){
  CV <- sd(fl, na.rm = T) / mean(fl, na.rm= T)
  return(CV)
}

# compute CV
CV <- Q_series %>% 
  summarise_if(is.numeric, ~compute_CV(.)) %>%  
  `rownames<-` ('CV')

#---- RBI --------
# function to compute RBI:
compute_RBI <- function(fl){
  fl_offset <- abs(fl - lag(fl))
  rbi <- sum(fl_offset, na.rm = T) / sum(fl, na.rm = T)
  return(rbi)
  }

# group by year
Q_series_annual <- Q_series %>% 
  group_by(img_yr)

# compute RBI
RBI <- Q_series_annual %>% 
  summarise_if(is.numeric, ~compute_RBI(.)) %>% # compute annual RBI
  ungroup %>% 
  summarise_if(is.numeric, ~mean(., na.rm = T)) %>%   # mean RBI across all years
  `rownames<-` ('RBI')

# --- DVIa  -----
Q_av <- Q_series %>% 
  summarise_if(is.numeric, ~mean(., na.rm = T))    # mean Q across all days

Q_monthly <- Q_series %>% # compute average monthly Q
  mutate(mth = as.character(month(date))) %>%   
  group_by(mth) %>%
  summarise_if(is.numeric, ~ifelse(sum(!is.na(.)) / 30 > completeness_thresh, mean(., na.rm = T), NA)) # only compute if month is sufficiently long

Q_monthly_min <- Q_monthly %>%  # driest month on average across all years
  ungroup %>% 
  summarise_if(is.numeric, ~min(., na.rm = T))  

Q_monthly_max <- Q_monthly %>% # wettest month on average across all years
  ungroup %>% 
  summarise_if(is.numeric, ~max(., na.rm = T))

DVIa <- ((Q_monthly_max - Q_monthly_min) / Q_av) %>% # work out DVIa
  `rownames<-` ('DVIa')


# ------DVIc-------

# function to compute DVIc:
compute_DVIc <- function(fl){
  qu <- quantile(fl, probs = c(0.05, 0.95), names = F, na.rm = T) # get 5th/95th quantiles of daily Q
  DVIc <- (qu[2] - qu[1]) / mean(fl, na.rm = T) # compute DVIc
  return(DVIc)
}

# compute DVIc
DVIc <- Q_series %>% 
  summarise_if(is.numeric, ~compute_DVIc(.)) %>%  # compute DVIc across whole record
  `rownames<-` ('DVIc')


```

## Compute central tendencies of discharge:

mean_Qd = mean daily discharge

mean_Qm = mean monthly discharge (months with insufficient data ignored)

mean_Qa = mean annual discharge (years with insufficient data ignored)

```{r Q_means}
mean_Qd <- Q_av %>% 
  `rownames<-` ('mean_Qd')

mean_Qm <- Q_monthly %>% 
  summarise_if(is.numeric, ~mean(., na.rm = T)) %>% # mean Q across all months with sufficient data
  `rownames<-` ('mean_Qm')

mean_Qa <- Q_series_annual %>% 
  summarise_if(is.numeric, ~ifelse(sum(!is.na(.)) / 365 > completeness_thresh, mean(., na.rm = T), NA)) %>%  # only compute if data is sufficiently long
  summarise_if(is.numeric, ~mean(., na.rm = T)) %>%  # mean Q across all years with sufficient data
  `rownames<-` ('mean_Qa')

```

## Write flow signatures and bulk metrics (central tendencies) to file:

A little reshaping first:

```{r write_outputs}
flow_sigs <- bind_rows(CV, RBI, DVIa, DVIc) %>% 
  t() %>% 
  as.data.frame() %>% 
  mutate(ohdb_id = row.names(.)) %>% 
  relocate(ohdb_id, .before = CV) 

Q_means <- bind_rows(mean_Qd, mean_Qm, mean_Qa) %>% 
  t() %>% 
  as.data.frame() %>% 
  mutate(ohdb_id = row.names(.)) %>% 
  relocate(ohdb_id, .before = mean_Qd) 

write.csv(flow_sigs, paste0(outdir, '/demo_flow_signatures.csv'), row.names = F)
write.csv(Q_means, paste0(outdir, '/demo_Q_means.csv'), row.names = F)
```

