---
title: "2_Paper_plots"
author: "Anya Leenman"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    self_contained: false
---
Code to generate the plots in ... submitted to ...

```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE)

library(raster)
library(lubridate)
library(sf)
library(dplyr)


rm(list = ls())


# Where are the supplementary data for this paper saved? 
# Available from Zenodo. DOI: 10.5281/zenodo.12193792
general_dir <- './data/SM_data/' 

# Where are water masks from Greenberg et al. 2024? This folder should contain the "SingleRivers" water masks, 
# available from: https://datadryad.org/stash/dataset/doi:10.5061/dryad.wm37pvmvf
water_mask_directory <- './data/mobility/mask/area/' 

# Where are the relevant hydrological data saved? 
# E.g. you could use the "tidyhydat" or "RivRetrieve" packages to download data.
# Daily hydrological data used in this study were saved as csv files with columns for date-stamps and discharge respectively. 
# Each filename corresponded to the flow-gauge id given in the 'ohdb_id' column of our supplementary data file, "ds02_Qvar_and_TR.csv". 

# Note that parts of the code for Figure 1 will not work unless you download the daily time-series data from the relevant agencies given in the manuscript.
hydro_timeseries_dir <- './data/hydrology/time_series/OHDB_Q/'



```

# Imports

Import data on mobility, channel character (planform, slope etc), sed flux, bulk discharge metrics, discharge variability.

Data on mobility and sediment flux from Greenberg and Ganti 2024, EPSL:
Data on bulk discharge metrics and discharge variability computed from previous script.

```{r imports}

df_joined <- read.csv(paste0(general_dir, '/ds02_Qvar_and_TR.csv'))

```

# Figure 1: 

Conceptual plot comparing raw flow data, flow duration curve + channel masks for a pair of rivers with similar mean discharge but different variability.

```{r concept_fig_data, eval = T}

# data prep ---- import data needed ------
# Koppen-geiger climate zones:
kgclim <- kgc::climatezones %>%
  mutate(KG = as.factor(substr(Cls, 1, 1))) %>%
  dplyr::select(-Cls) %>%
  dplyr::rename(x = Lon, y = Lat, z = KG) %>%
  relocate(x, .before = y) 

# Convert factor to integer codes
kgclim$z <- as.integer(kgclim$z)

# Create raster from dataframe
kg_rast <- rasterFromXYZ(kgclim)

# Set raster to categorical 
kg_rast <- ratify(kg_rast)

# Assign WGS CRS to raster
crs(kg_rast) <- CRS("+proj=longlat +datum=WGS84 +no_defs")

# Define Winkel II proj string
winkel_proj <- st_crs('ESRI:54019')$proj4string

# Reproject to Winkel II 
kg_rast_winkel <- projectRaster(kg_rast, crs = winkel_proj, method = "ngb")

# convert to SpatRaster
kg_rast_winkel <- terra::rast(kg_rast_winkel)


#-----
# reach locations as sf:
dfj_sf <- df_joined %>% 
  st_as_sf(coords = c(x = 'ohdb_longitude', y = 'ohdb_latitude'), crs = 4326) %>% 
  select(River, ohdb_id, notes, RBI)

dfj_sf_wink <- dfj_sf %>% 
  st_transform(crs = st_crs('ESRI:54019'))

# flow data for all sites - to generate panel with normalised FDCs 
flowdat_long <- lapply(1:nrow(df_joined), FUN = function(i){
  read.csv(paste0(hydro_timeseries_dir, df_joined$ohdb_id[i], '.csv')) %>% 
    mutate(ohdb_id = df_joined$ohdb_id[i]) %>% # unique ID for rows corresponding to each station
    mutate(Q_norm = Q / max(Q, na.rm = T)) %>% # normalised Q
    select(-Q, -date) %>% 
    arrange(desc(Q_norm)) %>% 
    filter(complete.cases(.)) %>% 
    mutate(rn = row_number()) %>% 
    mutate(perc_eq_ex = rn/max(rn) * 100) %>% # work out %time > Q_norm
    select(-rn)
}) %>% bind_rows() 

# ---sites of interest----
soi <- c(
  'Mortes_SantoAntonioDoLeverger', # low rbi, mean Q = 878.206986
  'Tisza_Vylok') # high rbi, mean Q = 189.744506
soi_codes <- c('MS', 'TV')
soi_names_pretty <- c('Mortes at Santo Antoniô do Leverger',   'Tisza at Vylok')

#----- flow records ------
soi_ohdb <- df_joined$ohdb_id[df_joined$River %in% soi] 
soi_flowdat <- lapply(1:length(soi), FUN = function(i){
  read.csv(paste0(hydro_timeseries_dir, soi_ohdb[i], '.csv')) %>% 
    mutate(ohdb_id = soi_ohdb[i]) %>% 
    mutate(date = ymd(date)) 
}) %>% bind_rows()

#------ water masks ------
require(terra)
masks <- lapply(1:length(soi), FUN = function(i){
  rastdir <- paste0(water_mask_directory, '/SingleRivers/RiverData/', 
                    df_joined$Type[df_joined$River == soi[i]], '/', soi[i])
  tifs <- list.files(rastdir, recursive = T, full.names = T, pattern = '.tif')
  tifs_aux <- list.files(rastdir, pattern = '.tif.aux', full.names = T, recursive = T)
  tifs <- base::setdiff(tifs, tifs_aux) # get rid of aux files from list of files to import
  st <- rast(tifs)
  
  # get timestamp from filename
  tvec_yr <- as.numeric(stringr::str_sub(tifs, -25, -22) )
  
  # error trap for date-stamp naming inconsistency in channel masks
  if(soi[i] == 'Tanana_NearHardingLake'){ 
    tvec_yr <- as.numeric(stringr::str_sub(tifs, -24, -21) )
  }
  
  # assign timestamps to raster stack time()
  time(st) <- tvec_yr 
  
  # zoom in if necessary so fig panels have similar scale
  if(soi[i] == 'Mortes_SantoAntonioDoLeverger'){
    st <- crop(st, y = ext(c(497850,  510000, 8631000, 8646000)))
  }
  
  return(st)
})
```

```{r colour_setup, eval = T}

# cols for plotting flow records: 
soi_cols <- c('#4575b4', '#d73027')

# cols for plotting rasters:
colvec <- scales::alpha(c('#a1dab4',
                          '#41b6c4', 
                          '#2c7fb8', 
                          '#253494'), 0.8)

# cols for plotting Koppen-Geiger climate zones:
kgcols <- c('#33a02c', # tropical
            '#fdc086', # dry
            '#b2df8a', # temperate
            '#ffff99', # continental
            '#beaed4') # polar

```

```{r conceptual_fig_base, fig.width = 7, fig.height = 8, dpi=300, eval = T}
par(oma = c(0, 0, 0 , 0),
    mai = c(0.5, 0.5, 0, 0),
    mgp = c(1.5, 0.5, 0))

layout(mat = matrix(1:(length(soi)*3), 
                    nrow = 3, 
                    ncol = length(soi)))  

for(i in 1:length(soi)){
  
  # row 1-------------------------
  
  if(i == 1){ 
    
    # site map - winkel II
    plot(kg_rast_winkel, 
         pax = list(side = NA),
         col = kgcols,
         mar = rep(0, 4),
         legend = F,
         xlim = c(-12000000, 15000000),
         ylim = c(-7000000, 9500000),
         box = F)
    
    # annotations
    par(xpd = NA)
    add_legend(x = -12000000, y = -5800000, 
               legend = c(LETTERS[1:5], soi_codes), 
               border = NA,
               fill = c(kgcols, NA, NA), 
               pch = c(rep(NA, 5), rep(25, 2)),
               col = NA,
               pt.bg = c(rep(NA, 5), soi_cols),
               horiz = T, 
               cex = 1.2,
               bty = 'n',
               x.intersp = 0.3)
    add_legend(x = -12000000, y = -100000,
               title = 'RBI',
               legend = signif(range(dfj_sf_wink %>% pull('RBI')), 2),
               pt.cex = range(dfj_sf_wink %>% pull('RBI')) * 7 + 1,
               pch = 19,
               col = scales::alpha('black', 0.5),
               bty = 'n')
    add_legend(x = -12250000, y = 12300000,
               legend = '(a)',
               text.font = 2,
               cex = 1.2,
               bty = 'n')
    par(xpd = F)
    
    # river gauge locations
    points(dfj_sf_wink %>% filter(!ohdb_id %in% soi_ohdb),
           cex = dfj_sf_wink %>% filter(!ohdb_id %in% soi_ohdb) %>% pull('RBI') * 7 + 1,
           pch = 19,
           col = scales::alpha('black', 0.5))
    
    points(dfj_sf_wink %>% filter(ohdb_id %in% soi_ohdb),
           cex = dfj_sf_wink %>% filter(ohdb_id %in% soi_ohdb) %>% pull('RBI') * 7 + 1,
           bg = soi_cols,
           pch = 25,
           col = 'white')
    
    
  } else if (i == 2){ # FDCs of all - normalised
    
    # blank axes
    plot(c(0, 100), range(flowdat_long$Q_norm), type = 'n',
         ylab = bquote('Normalized discharge,' ~italic('Q / Q max')),
         xlab = '% Time equalled or exceeded',
         log = 'y',
         ylim = c(0.003, 1.1))
    
    # fdc lines
    for(j in 1:nrow(df_joined)){
      fd <- flowdat_long %>% filter(ohdb_id == df_joined$ohdb_id[j])
      lines(fd$perc_eq_ex,
            fd$Q_norm,
            col = scales::alpha('darkgray', 0.5))
    }
    
    # highlights sites of interest:
    leg_pos <- c('topright', 'bottomleft')
    for(j in 1:length(soi)){
      fd <- flowdat_long %>% filter(ohdb_id == soi_ohdb[j])
      lines(fd$perc_eq_ex,
            fd$Q_norm,
            lwd = 2,
            col = soi_cols[j])
      # annotations:
      legend(leg_pos[j], soi_names_pretty[j], bty = 'n', text.col = soi_cols[j])
    }
    
    # subfig label
    par(xpd = NA)
    legend(x = -22, y = 1.7,
           legend = '(b)',
           text.font = 2,
           cex = 1.2,
           bty = 'n')
    par(xpd = T)
  }
  
  # row 2: ----plot raw flow - note time crop to 'zoom' on same period----
  
  Q <- soi_flowdat %>% 
    filter(ohdb_id == soi_ohdb[i]) %>% 
    filter(date >= ymd('1990-01-01') & date <= ymd('2010-12-31')) 
  
  plot(Q$date, Q$Q,
       xlab = 'Date', 
       ylab = bquote('Discharge,' ~ italic(Q) ~ ' (m'^3*s^-1*')'), 
       type = 'l',
       cex.lab = 1.2,
       ylim = range(soi_flowdat$Q, na.rm = T))
  
  # annotation
  legend('topright', soi_names_pretty[i], text.col = soi_cols[i], bty = 'n')
  
  # subfig label
  par(xpd = NA)
  legend(x = ymd('1985-06-01'), y = 4000,
         legend = paste0('(', letters[i+2], ')'),
         text.font = 2,
         cex = 1.2,
         bty = 'n')
  par(xpd = T)
  
  
  # row 3---plot mobility masks----
  
  # annotations (technically attached to row above!):
  par(xpd = NA)
  legend(
    'bottomleft',
    legend = c(
      bquote(RBI == ~ .(signif(df_joined$RBI[df_joined$ohdb_id == soi_ohdb[i]], 2))),
      bquote(DVI[a] == ~ .(signif(df_joined$DVIa[df_joined$ohdb_id == soi_ohdb[i]], 2))),
      bquote(DVI[c] == ~ .(signif(df_joined$DVIc[df_joined$ohdb_id == soi_ohdb[i]], 2))),
      bquote(bar(Q) == ~ .(signif(df_joined$mean_Qd[df_joined$ohdb_id == soi_ohdb[i]], 2)) ~ m^3 * s^-1),
      bquote(bar(B) == ~ .(signif(df_joined$Width..m.[df_joined$ohdb_id == soi_ohdb[i]], 2)) ~ m),
      bquote('T'[R] == ~ .(signif(df_joined$TR[df_joined$ohdb_id == soi_ohdb[i]], 2)) ~ yr)),
    inset = c(-0.1, -0.7),
    bty = 'n',
    cex = 1.2,
    y.intersp = 1.2
  )
  legend(x = ymd('1985-06-01'), y = -800,
         legend = paste0('(', letters[i+4], ')'),
         text.font = 2,
         cex = 1.2,
         bty = 'n')
  par(xpd = F)
  
  # vector of years to plot
  yrs <- c(1990, 2000, 2010, 2020) 
  
  # start index - correction to make sure end of colour scale used in plot
  si <- length(colvec) + 1 - length(yrs)
  
  # plot 1 + add next decades
  plot(masks[[i]][[time(masks[[i]]) == yrs [1]]],
       col = c(NA, colvec[si]),
       legend = F,
       axes = F,
       mar = rep(0, 4),
       buffer = F,
       clip = T) # first raster
  pu <- par('usr') # save axis lims to object
  sbar(2000, xy="right", divs=2, cex=1, ticks=TRUE, label = ('2 km')) # add scalebar
  
  
  for(j in 2:length(yrs)){ # superimpose next layers - cleaner way?
    plot(masks[[i]][[time(masks[[i]]) == yrs [j]]], 
         col = c(NA, colvec[si-1 + j]),
         axes = F,
         legend = F,
         mar = rep(0, 4),
         buffer = F,
         clip = T,
         add = T) # additional raster
  }
}

par(xpd = NA)
xadj <- -0.35
yadj <- 0.15 # tweak as necessary
add_legend(
  x = pu[1] + xadj * (pu[2] - pu[1]), 
  y = pu[3] + yadj * (pu[4] - pu[3]),
  legend = yrs,
  fill = colvec[si:length(colvec)],
  border = NA,
  horiz = T, 
  bty = 'n',
  cex = 1.2)
par(xpd = F)
```


# Figure 2: 

Channel mobility T_{R} vs bulk metrics of flow and topography. 

Prep cols:

```{r f2cols}
require(scales)
pcols <- RColorBrewer::brewer.pal(3, 'Set1')[2:3]

```

Make plot:

```{r bulkmetrics_v_TR, fig.height=6, fig.width=6, dpi=300, eval = T}
# plotting params
lposy <- c(161, 148, 135)
xadj <- -0.2

par(mfrow = c(2,2), mai = c(0.5, 0.5, 0.7, 0), mgp=c(1.8,0.5,0))

# variables to include
bulkvars <- c('mean_Qd', 'CV', 'slope..m.m', 'Bed_Qs_norm')
xlabs <- c(bquote(italic(bar(Q)) ~ '('~m^3 * s^-1~')'),
           bquote('Coefficient of variation (CV) of' ~ italic(Q)),
           bquote('Channel slope,' ~ italic(S) ~ ' (-)'),
           bquote('Sediment concentration,' ~ italic(Q[s]^'*') ~ ' (-)')
           )

# subdivide by n channelthreads
div_var <- 'Channel.Count.Index....'
div_thresh <- 1
group_labels <- c('All', 'Single-thread', 'Multi-thread')

# subset data
sub1 <- df_joined %>% filter(!!as.symbol(div_var) <= !!div_thresh)
sub2 <- df_joined %>% filter(!!as.symbol(div_var) > !!div_thresh) 
subs <- list(sub1, sub2)

# plot
for(i in 1:length(bulkvars)){
  bvar <- colnames(subset(df_joined, select = bulkvars))[i]
  
  # set plot axes
  plot(range(df_joined[bvar], na.rm = T), range(df_joined$TR),
       xlab = xlabs[i], 
       ylab = bquote('Reworking timescale,' ~ italic('T'[R]) ~ ' (yr)'), 
       type = 'n', 
       log = '')  # if log-transform needed
  pu <- par('usr') # save axis lims to object
  
  # fit exponential model to full dataset:
  x = df_joined[bvar] %>% pull() %>% replace(., .<=0.0, NA) # substitute zeros or negatives with NA
  y = df_joined$TR
  
  mod <- lm(log10(y) ~ log10(x))
  r2a <- summary(mod)$adj.r.squared
  pv <- signif(summary(mod)$coefficients[2,4], 1)
  
  # plot model + conflims
  newx <- seq(min(x, na.rm = T), max(x, na.rm = T), length.out = 999)
  mod.preds <- 10^(predict(mod, interval = 'confidence', new = list(x = newx), level = 0.95))
  polygon(c(newx, rev(newx)), # add conflims to plot
          c(mod.preds[, 2], rev(mod.preds[, 3])),
          col = 'gray85',
          border = NA)
  lines(newx, mod.preds[,1], col = 'black') # add model to plot
  
  # annotations
  par(xpd = NA)
  legend(x = pu[1] + xadj * (pu[2] - pu[1]),
         y = lposy[1],
         legend = bquote(.(group_labels[1]) ~
                           italic('r'^2) ~ '= ' ~ .(signif(r2a, 2)) ~
                           italic('p ') ~ .(ifelse(pv >= 0.001, paste0('= ', pv), '< 0.001'))),         
         bg = NA,
         bty = 'n')
  legend('topright', legend = paste0('(', letters[i], ')'), text.font = 2, bty = 'n')
  par(xpd = F)
  
  for(j in 1:length(subs)){
    
    x <- subs[[j]][bvar] %>% pull() %>% replace(., .<=0.0, NA) # substitute zeros or negatives with NA
    y <- subs[[j]]$TR
    
    # fit exponential model:
    mod <- lm(log10(y) ~ log10(x))
    r2a <- summary(mod)$adj.r.squared
    pv <- signif(summary(mod)$coefficients[2,4], 1)
    
    # plot points
    points(x, y, col = alpha(pcols[j], 0.7), pch = 16)
    
    # plot model - subsets
    newx = seq(min(x, na.rm = T), max(x, na.rm = T), length.out = 999)
    mod.preds <- 10^(predict(mod,list(x = newx)))
    lines(newx, mod.preds, col = pcols[j])
    
    # annotations
    par(xpd = NA)
    legend(x = pu[1] + xadj * (pu[2] - pu[1]),
           y = lposy[j+1],
           col = pcols[j],
           legend = bquote(.(group_labels[j+1]) ~
                             italic('r'^2) ~ '= ' ~ .(signif(r2a, 2)) ~
                             italic('p ') ~ .(ifelse(pv >= 0.001, paste0('= ', pv), '< 0.001'))),
           bg = NA,
           text.col = pcols[j],
           bty = 'n')
    par(xpd = F)
  }
}

```

# Figure 3: 

Channel mobility T_{R} vs discharge variability. 

```{r Qvar_v_TR, fig.height=2.5, fig.width=6, dpi=300, eval = T}
# plotting params
par(mfrow = c(1,3), oma = c(0, 0, 0, 0.7), mai = c(0.4, 0.4, 0.6, 0.07), mgp=c(1.8,0.5,0))

# variables to include
Qvars <- c('RBI', 'DVIa', 'DVIc')
xlabs <- c(bquote('Richards-Baker Index,' ~ italic(RBI) ~ '(-)'),
           bquote('Annual discharge var.,' ~ italic(DVI[a]) ~ '(-)'),
           bquote('Cumulative discharge var.,' ~ italic(DVI[c]) ~ '(-)'))

# plot
for(i in 1:length(Qvars)){
  bvar <- colnames(subset(df_joined, select = Qvars))[i]
  
  # set plot axes
  par(xpd = NA)
  plot(range(df_joined[bvar], na.rm = T), range(df_joined$TR),
       xlab = xlabs[i], 
       ylab = bquote('Reworking timescale,' ~ italic('T'[R]) ~ ' (yr)'), 
       type = 'n', 
       log = '')  # if log-transform needed
  par(xpd = F)
  pu <- par('usr') # save axis lims to object
  
  # fit exponential model to full dataset:
  x = df_joined[bvar] %>% pull() %>% replace(., .<=0.0, NA) # substitute zeros or negatives with NA
  y = df_joined$TR
  
  mod.exp <- lm(log10(y) ~ log10(x))
  r2a <- summary(mod.exp)$adj.r.squared
  pv <- signif(summary(mod.exp)$coefficients[2,4], 1)
  
  # plot model + conflims
  newx <- seq(min(x, na.rm = T), max(x, na.rm = T), length.out = 999)
  mod.exp.preds <- 10^(predict(mod.exp, interval = 'confidence', new = list(x = newx), level = 0.95))
  polygon(c(newx, rev(newx)), # add conflims to plot
          c(mod.exp.preds[, 2], rev(mod.exp.preds[, 3])),
          col = 'gray85',
          border = NA)
  lines(newx, mod.exp.preds[,1], col = 'black') # add model to plot
  
  # annotations
  par(xpd = NA)
  legend(x = pu[1] + xadj * (pu[2] - pu[1]),
         y = lposy[1],
         legend = bquote(.(group_labels[1]) ~
                           italic('r'^2) ~ '= ' ~ .(signif(r2a, 2)) ~
                           italic('p ') ~ .(ifelse(pv >= 0.001, paste0('= ', pv), '< 0.001'))),      
         bg = NA,
         bty = 'n')
  legend('topright', legend = paste0('(', letters[i], ')'), text.font = 2, bty = 'n')
  par(xpd = F)
  
  for(j in 1:length(subs)){
    
    x <- subs[[j]][bvar] %>% pull() %>% replace(., .<=0.0, NA) # substitute zeros or negatives with NA
    y <- subs[[j]]$TR
    
    # fit exponential model:
    mod.exp <- lm(log10(y) ~ log10(x))
    r2a <- summary(mod.exp)$adj.r.squared
    pv <- signif(summary(mod.exp)$coefficients[2,4], 1)
    
    # plot points
    points(x, y, col = alpha(pcols[j], 0.7), pch = 16)
    
    # plot model - subsets
    newx = seq(min(x, na.rm = T), max(x, na.rm = T), length.out = 999)
    mod.exp.preds <- 10^(predict(mod.exp,list(x = newx)))
    lines(newx, mod.exp.preds, col = pcols[j])
    
    # annotations
    par(xpd = NA)
    
    legend(x = pu[1] + xadj * (pu[2] - pu[1]),
           y = lposy[j+1],
           col = pcols[j],
           legend = bquote(.(group_labels[j+1]) ~
                             italic('r'^2) ~ '= ' ~ .(signif(r2a, 2)) ~
                             italic('p ') ~ .(ifelse(pv >= 0.001, paste0('= ', pv), '< 0.001'))),
           bg = NA,
           text.col = pcols[j],
           bty = 'n')
    par(xpd = F)
  }
}
```


